\LARGE{\bf{Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})}}\footnotemark

### 2014-05-23 manuscript.tex
edits
edits
continued final edits

Remaining explanations for differential frequencies of the \emph{Hopscotch} among teosinte populations include both genetic drift and natural selection.  Previous studies using both SSRs and genome-wide SNP data have found evidence for a population bottleneck in the San Lorenzo population \citep{Hufford2010, Pyhajarvi2013}, and the lower levels of sequence diversity in this population in the 5' UTR (Region 1) coupled with more positive values of Tajima's D are consistent with these earlier findings.  Such population bottlenecks can exaggerate the effects of genetic drift through which the \emph{Hopscotch} allele may have risen to high frequency entirely by chance. A bottleneck in San Lorenzo, however, does not explain the high frequency of the \emph{Hopscotch} in multiple populations in the Jalisco cluster.  Moreover, available information on diversity and population structure among Jaliscan populations \citep{Hufford2010, Pyhajarvi2013} is not suggestive of recent colonization or other demographic events that would predict a high frequency of the allele across populations.  Finally, diversity values in the 5' UTR of \emph{tb1} are suggestive of natural selection acting upon the gene in natural populations of \emph{parviglumis}.  Overall nucleotide diversity is \jri{X\%} less than seen in the sequences from the 66kb upstream regions, and Tajima's D is considerably lower and consistently negative.  Though not definitive, these results are consistent with the action of selection on the upstream region of \emph{tb1}, perhaps suggesting an ecological role for the gene in \emph{parviglumis}. \jri{whoa there Dr. Gillespie! i'm not so sure of this. first illumina data are wonky and the TajD from Hufford2012b drops all singletons. if we are to compare to other data it should be to Tiffin and Moeller. do we know the Hop genotype for sequenced lines? can we separate the sequences into hop/no hop and look for differences? it wasn't until we did this that gt1 stuff really popped out.} \lev{we should know for some of them, i will check} \mbh{I've added a few sentences on selection.  Do we still want to compare sequences with and without \emph{Hopscotch}?  I agree its a good idea and could end up being really interesting.  Perhaps something we could look at after submission and incorporate during revisions?} \jri{ sure, we can look after submission, but i want to do that before publication. to some degree i'd feel better having stuff done before submission especially if we want to argue selection, but if you guys want to push to submit, this and github stuff can wait until revision}

Significant effects of the \emph{Hopscotch} insertion on lateral branch length, number of cupules, and tillering index in domesticated maize have been well documented \citep{Studer2011}.  \citet{Weber2007}  described significant phenotypic associations between markers in and around \emph{tb1} and lateral branch length and female ear length in a sample from 74 natural populations of \emph{parviglumis} \citep{Weber2007}; however, these data did not include markers from the \emph{Hopscotch} region 66kb upstream of \emph{tb1}. Our study is the first to explicitly examine the phenotypic effects of the \emph{Hopscotch} insertion across a wide collection of individuals sampled from natural populations of teosinte. We have found no significant effect of the \emph{Hopscotch} insertion on tillering index or tiller number, a result that is discordant with its clear phenotypic effects in maize. One interpretation of this result would be that the \emph{Hopscotch} controls tillering in maize \citep{Studer2011}, but tillering in teosinte is affected by variation at other loci. Consistent with this interpretation, \emph{tb1} is thought to be part of a complex pathway controlling branching, tillering and other phenotypic traits \citep{KebromBrutnell2007, Clark2006}. A recent study by \citet{StuderDoebley2012} examined variation across traits in a three-taxa allelic series at the \emph{tb1} locus. \citet{StuderDoebley2012} introgressed nine unique teosinte \emph{tb1} segments (one from \emph{Zea diploperennis}, and four each from \emph{mexicana} and \emph{parviglumis}) into an inbred maize background and investigated their phenotypic effects. Phenotypes were shown to cluster by taxon, indicating \emph{tb1} may underlie morphological diversification of \emph{Zea}. Additional analysis in \citet{StuderDoebley2012} suggested tillering index was controlled both by \emph{tb1} and loci elsewhere in the genome.  Clues to the identity of these loci may be found in QTL studies that have identified loci controlling branching architecture (\emph{e.g.,} \citealt{DoebleyStec1991, DoebleyStec1993}). Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012},  and both \emph{tru1} and \emph{ter1} affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tru1}, for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number; \citealt{Li2012}). Variation in these additional loci may have affected tillering in our collections and contributed to the lack of correlation we see between \emph{Hopscotch} genotype and tillering.  \jri{worth also mentioning possibly crowding or other non-normal factors of our greenhouse?}

>>>>

Remaining explanations for differential frequencies of the \emph{Hopscotch} among teosinte populations include both genetic drift and natural selection.  Previous studies using both SSRs and genome-wide SNP data have found evidence for a population bottleneck in the San Lorenzo population \citep{Hufford2010, Pyhajarvi2013}, and the lower levels of sequence diversity in this population in the 5' UTR (Region 1) coupled with more positive values of Tajima's D are consistent with these earlier findings.  Such population bottlenecks can exaggerate the effects of genetic drift through which the \emph{Hopscotch} allele may have risen to high frequency entirely by chance. A bottleneck in San Lorenzo, however, does not explain the high frequency of the \emph{Hopscotch} in multiple populations in the Jalisco cluster.  Moreover, available information on diversity and population structure among Jaliscan populations \citep{Hufford2010, Pyhajarvi2013} is not suggestive of recent colonization or other demographic events that would predict a high frequency of the allele across populations.  Finally, diversity values in the 5' UTR of \emph{tb1} are suggestive of natural selection acting upon the gene in natural populations of \emph{parviglumis}.  Overall nucleotide diversity is 76\% less than seen in the sequences from the 66kb upstream region, and Tajima's D is considerably lower and consistently negative.  In fact, values of Tajima's D in the 5' UTR are toward the extreme negative end of the distribution of this statistic previously calculated across loci sequenced in \emph{parviglumis} \citep{Wright27052005, Moeller01072007}.  Though not definitive, these results are consistent with the action of selection on the upstream region of \emph{tb1}, perhaps suggesting an ecological role for the gene in \emph{parviglumis}.

Significant effects of the \emph{Hopscotch} insertion on lateral branch length, number of cupules, and tillering index in domesticated maize have been well documented \citep{Studer2011}.  \citet{Weber2007}  described significant phenotypic associations between markers in and around \emph{tb1} and lateral branch length and female ear length in a sample from 74 natural populations of \emph{parviglumis} \citep{Weber2007}; however, these data did not include markers from the \emph{Hopscotch} region 66kb upstream of \emph{tb1}. Our study is the first to explicitly examine the phenotypic effects of the \emph{Hopscotch} insertion across a wide collection of individuals sampled from natural populations of teosinte. We have found no significant effect of the \emph{Hopscotch} insertion on tillering index or tiller number, a result that is discordant with its clear phenotypic effects in maize. One interpretation of this result would be that the \emph{Hopscotch} controls tillering in maize \citep{Studer2011}, but tillering in teosinte is affected by variation at other loci. Consistent with this interpretation, \emph{tb1} is thought to be part of a complex pathway controlling branching, tillering and other phenotypic traits \citep{KebromBrutnell2007, Clark2006}. A recent study by \citet{StuderDoebley2012} examined variation across traits in a three-taxa allelic series at the \emph{tb1} locus. \citet{StuderDoebley2012} introgressed nine unique teosinte \emph{tb1} segments (one from \emph{Zea diploperennis}, and four each from \emph{mexicana} and \emph{parviglumis}) into an inbred maize background and investigated their phenotypic effects. Phenotypes were shown to cluster by taxon, indicating \emph{tb1} may underlie morphological diversification of \emph{Zea}. Additional analysis in \citet{StuderDoebley2012} suggested tillering index was controlled both by \emph{tb1} and loci elsewhere in the genome.  Clues to the identity of these loci may be found in QTL studies that have identified loci controlling branching architecture (\emph{e.g.,} \citealt{DoebleyStec1991, DoebleyStec1993}). Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012},  and both \emph{tru1} and \emph{ter1} affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tru1}, for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number; \citealt{Li2012}). Variation in these additional loci may have affected tillering in our collections and contributed to the lack of correlation we see between \emph{Hopscotch} genotype and tillering. Finally, although photoperiod for Phenotyping 2 reasonably approximated that of the normal \emph{parviglumis} growing season, greenhouse-specific environmental conditions (plant density, light regime, etc...) may have contributed to tillering responses different from those found in nature, obscuring the effect of the \emph{Hopscotch} insertion on variation.
